Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-07-30 surfaces-write-up/surface-write-up.tex
surfaces: reword 1pt to eff, add idea of pf to eff

\ssec{Main Results}
\todo{Aaron: Some of these theorem statements have been copied from older versions of the statements
and not updated to reflect the difference. I agree we should use Theorem-restate, but we should make
sure to copy the ones stated below, and not copy these ones. Robin: thm-restate only allows you to restate earlier theorems and not later ones, so the main thms must appear numbered as 1.-}
\begin{thm}
\label{thm:proj-effective-intro}
Let $X = \bp^m$ with $\{x_0, \ldots, x_m\}$ a coordinate system for
$\bp^m$. Let $D = \sum_{i = 1}^{n} \alpha_i C_i$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$ and each $C_i$ a hypersurface of
$\bp^m$, be an effective divisor on $\bp^m$.

@@

\begin{thm}
\label{thm:proj-one-point}
Let $\bp^m \cong \proj \bk [x_0, \ldots, x_m].$ Let $D = \alpha H$, with $\alpha = \frac{p}{q} \in \bq$
and $H := V(x_k)$ a hyperplane of $\bp^m$.
Let
\[
	0 = \frac{c_0}{d_0} <
	\frac{c_1}{d_1} < \ldots < \frac{c_r}{d_r} = \frac{p}{q}
\]

@@

\noindent
has a minimal presentation consisting of the $\sum_{i = 0}^{r}
{{m + c_i - 1} \choose {c_i}}$ generators $F_i^{\vec{v}} := \frac{u^{d_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $1 \leq i \leq r$, $\vec{v} \in \bz_{\geq 0}^{m + 1}$
with $\deg \vec v = c_i$. Furthermore, it has
relations of the following two types:
\begin{enumerate}
	\item For each $(i, j)$ with $j \geq i + 2$ and each $\vec{v} \in \mss_i,
\vec{w} \in \mss_j$, there is a relation of the form either
\begin{flalign*}
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})
	^{g_{\vec{y}}} &(i < h_{i, j} < j) \\
	\text{or} \\
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})^{g_{\vec{y}}}
	\cdot \prod_{ \vec{z} \in
	\mss_{h_{i, j} + 1}} (F_{k_{i, j} + 1}^{\vec{z}})^{g'_{\vec{z}}}
	&(i < h_{i, j} < h_{i, j} + 1 < j).
\end{flalign*}
	\item For each $(i, j)$ with
$j = i$ or $j = i + 1$ and each $\vec{v} \in \mss_i, \vec{w} \in
\mss_j$ with $\vec{v} \not\prec \vec{w}$ (see Definition
~\ref{defn:vec-order}) there is a relation of the form
\begin{align*}
	&L_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- F_i^{\vec{y}} F_j^{\vec{z}} \\
\end{align*}

@@

\noindent
where $\vec{y}$ and $\vec{z}$ are the unique
vectors in $\mss_i$ and $\mss_j$, respectively, such that $\vec{y}
+ \vec{z} = \vec{v} + \vec{w}$ and $\vec{y} \prec \vec{z}$.
\end{thm}
\todo{Write idea of proof}
\begin{proof}
We will first check that $F^{\vec v}_i$ determine a system of minimal
generators of $R_D$. We know $\left\{F_{1}^{\vec{v}} : \deg \vec{v} =
\left\lfloor
\frac{p}{q} \right\rfloor \right\}$ are minimal generators in
degree $1$. Now let $i > 1$. Suppose $F_i^{\vec{v}}$ were not
minimal for some $\vec{v} \in \mss_i$. Then $F_i^{\vec{v}} =
F_{s}^{\vec{y}} F_{t}^{\vec{z}}$ for some $s, t \in \bn$ and
$\vec{y}, \vec{z} \in \bz_{ \geq 0}^{m + 1}$. In particular, we see
that 
\begin{align*}
	&d_i = s + t \\
	&c_i = \deg \vec{v} = \deg \vec{y} + \deg \vec{z}
\end{align*}

@@

\begin{rem}
\label{rem:proj-one-point-ind}
The method of proof for Theorem ~\ref{thm:proj-one-point}
can be immediately generalized to work inductively. If $D = D' + \alpha H$
for some hyperplane $H$, then the proof of Theorem ~\ref{thm:proj-one-point}, mutatis mutandis, extends to show
the canonical ring $R_D$ is generated over $R_D'$ by the same
generators and with the same relations given in Theorem
~\ref{thm:proj-one-point}.
\end{rem}
\todo{Perhaps we should consider restating the above theorem in this form, and then say for simplicity we prove it when the base is 0?}
\begin{thm}
\label{thm:proj-effective}
Let $[x_0, \ldots, x_m]$ be a projective coordinate system for
$\bp^m$. Let $D = \sum_{i = 1}^{n} \alpha_i D_i \in \bq \otimes \bp^m$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$.

@@

\begin{proof}
We derive the result from the following inductive argument.
If $n = 0$, i.e.~ $D = 0$, then we are done.
Now, we look to inductively add hypersurfaces.
If $D' = \sum_{i = 0}^{n} \alpha_i D_i$ is some $\bq$-divisor
on $\bp^m$, then we look at $D = D' + C$ where $C$ is a
hypersurface of degree $\pdeg$. \todo{Aaron: rewrite this to make sense} Now use the Veronese mapping
to view $C$ as a hypersurface on $\bp^{\binom{{m + \pdeg}}{ \pdeg} -1 }$
and the image of $C$ as the intersection of a Veronese variety
with a hyperplane. Then we can reduce inductively adding hypersurfaces
to the case of adding hyperplanes. However, this is accomplished
by the method used in Theorem ~\ref{thm:proj-one-point} in an
inductive manner as mentioned in Remark ~\ref{rem:proj-one-point-ind}.
\end{proof}

>>>>

\ssec{Main Results}
\begin{thm}
\label{thm:proj-effective-intro}
Let $X = \bp^m$ with $\{x_0, \ldots, x_m\}$ a coordinate system for
$\bp^m$. Let $D = \sum_{i = 1}^{n} \alpha_i C_i$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$ and each $C_i$ a hypersurface of
$\bp^m$, be an effective divisor on $\bp^m$.

@@

\begin{thm}
\label{thm:proj-one-point}
Let $\bp^m \cong \proj \bk [x_0, \ldots, x_m].$ Let $D' \in \bq
\otimes \div \bp^m$ and $D = D' + \alpha H$, with $\alpha =
\frac{p}{q} \in \bq$ and $H := V(x_k)$ a hyperplane of $\bp^m$.
Let
\[
	0 = \frac{c_0}{d_0} <
	\frac{c_1}{d_1} < \ldots < \frac{c_r}{d_r} = \frac{p}{q}
\]

@@

\noindent
has a minimal presentation over $R_{D'}$ consisting of the $\sum_{i = 0}^{r}
{{m + c_i - 1} \choose {c_i}}$ generators $F_i^{\vec{v}} := \frac{u^{d_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $1 \leq i \leq r$, $\vec{v} \in \bz_{\geq 0}^{m + 1}$
with $\deg \vec v = c_i$. Furthermore, it has
relations generated \todo{over $R_{D'}$?} by the following two types:
\begin{enumerate}
	\item For each $(i, j)$ with $j \geq i + 2$ and each $\vec{v} \in \mss_i,
\vec{w} \in \mss_j$, there is a relation of the form either
\begin{flalign*}
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})
	^{g_{\vec{y}}} &(i < h_{i, j} < j) \\
	\text{or} \\
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})^{g_{\vec{y}}}
	\cdot \prod_{ \vec{z} \in
	\mss_{h_{i, j} + 1}} (F_{k_{i, j} + 1}^{\vec{z}})^{g'_{\vec{z}}}
	&(i < h_{i, j} < h_{i, j} + 1 < j).
\end{flalign*}
	\item For each $(i, j)$ with
$j = i$ or $j = i + 1$ and each $\vec{v} \in \mss_i, \vec{w} \in
\mss_j$ with $\vec{v} \not\prec \vec{w}$ (see Definition
~\ref{defn:vec-order}) there is a relation of the form
\begin{align*}
	&L_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- F_i^{\vec{y}} F_j^{\vec{z}} \\
\end{align*}

@@

\noindent
where $\vec{y}$ and $\vec{z}$ are the unique
vectors in $\mss_i$ and $\mss_j$, respectively, such that $\vec{y}
+ \vec{z} = \vec{v} + \vec{w}$ and $\vec{y} \prec \vec{z}$.
\end{thm}

\sssec*{Idea of Proof:}
The proof follows in four steps. First, we demonstrate that
the $F_{i}^{\vec{v}}$'s form a set of minimal generators of $R_D$
over $R_{D'}$ by leveraging the fact that $F_{i}^{\vec{v}}$ has
``maximal pole order'' at some coordinate among rational functions
in degree $i$. Then, the relations $G_{i, j}^{\vec{v}, \vec{w}}$
follow from the demonstration that the $F_{i}^{\vec{v}}$'s generate
all of $R_D$ over $R_{D'}$. We derive the relations
$L_{i, j}^{\vec{v}, \vec{w}}$ by considering when products of
generators in neighboring degrees are equal. Finally, we demonstrate
that $G_{i, j}^{\vec{v}, \vec{w}}$'s and $L_{i, j}^{\vec{v}, \vec{w}}$'s
generate all of the relations by using them to reduce elements of
$R_D$ to a canonical form.

\begin{proof}
For simplicity, we give a proof for when $D' = 0$.
We will first check that $F^{\vec v}_i$ determine a system of minimal
generators of $R_D$. We know $\left\{F_{1}^{\vec{v}} : \deg \vec{v} =
\left\lfloor
\frac{p}{q} \right\rfloor \right\}$ are minimal generators in
degree $1$. Now let $i > 1$. Suppose $F_i^{\vec{v}}$ were not
minimal for some $\vec{v} \in \mss_i$. Then $F_i^{\vec{v}} =
F_{s}^{\vec{y}} F_{t}^{\vec{z}}$ for some $s, t \in \bn$ and
$\vec{y}, \vec{z} \in \bz_{ \geq 0}^{m + 1}$. In particular, we see
that 
\begin{align*}
	&d_i = s + t \\
	&c_i = \deg \vec{v} = \deg \vec{y} + \deg \vec{z}
\end{align*}

@@

%\begin{rem}
%\label{rem:proj-one-point-ind}
%The method of proof for Theorem ~\ref{thm:proj-one-point}
%can be immediately generalized to work inductively. If $D = D' + \alpha H$
%for some hyperplane $H$, then the proof of Theorem ~\ref{thm:proj-one-point}, mutatis mutandis, extends to show
%the canonical ring $R_D$ is generated over $R_D'$ by the same
%generators and with the same relations given in Theorem
%~\ref{thm:proj-one-point}.
%\end{rem}

\begin{thm}
\label{thm:proj-effective}
Let $[x_0, \ldots, x_m]$ be a projective coordinate system for
$\bp^m$. Let $D = \sum_{i = 1}^{n} \alpha_i D_i \in \bq \otimes \bp^m$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$.

@@

\begin{proof}
We derive the result from the following inductive argument.
If $n = 0$, i.e.~ $D = 0$, then we are done.
Now, we look to inductively add hypersurfaces.
If $D' = \sum_{i = 0}^{n} \alpha_i D_i$ is some $\bq$-divisor
on $\bp^m$, then we look at $R_D$ where $D = D' + C$ for some
degree $\pdeg$ hypersurface $C$. Now use the Veronese mapping
to view $C$ as a hypersurface on $\bp^{\binom{{m + \pdeg}}{\pdeg} - 1}$
and the image of $C$ as the intersection of a Veronese variety
with a hyperplane. Then we can reduce the process of inductively
adding hypersurfaces to when the hypersurfaces are hyperplanes. However, this is accomplished
by the method used in Theorem ~\ref{thm:proj-one-point} in an
inductive manner as mentioned in Remark ~\ref{rem:proj-one-point-ind}.
\end{proof}
